On the number of tiles touched by a\\ line segment on a rectangular grid

### 2021-07-31 paper.tex
"Make room" in the paper structure for the average-number-of-tiles problem
This involves new section/subsection structure and rewriting the introduction.

I also include some ideas to be developed in the new section for the average number of tiles

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gices infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\mss}{M}

@@

\title{
On the maximum number of tiles touched\\
by a line segment on a rectangular grid
}

% Authors with different affiliations: from https://tex.stackexchange.com/a/370813/90222
\author[1]{Luis Mendo}
\author[2]{**Name**}

@@

Given $a, b \in \mathbb R^+$, consider a grid on $\mathbb R^2$ formed by rectangular \emph{tiles} of width $a$ and height $b$. A line segment of length $\len \in \mathbb R^+$ is located on the plane with arbitrary position and orientation. The segment is said to \emph{touch} a tile if it intersects its interior.\footnote{
The definition uses the interior of the tile, excluding the border, to avoid uninteresting results such as a ``zero-length'' segment touching (a vertex of) $4$ tiles.}
What is the maximum number of tiles that the segment can touch?

As an example, consider $a=1.35$, $b=1$. A segment of unit length can be placed as shown in Figure~\ref{fig: examples} (left) to touch $3$ tiles. In fact, the solution for $\len=1$ is $3$ tiles. The figure also illustrates that the solution for length $2.4$ is $5$ (center), and for $4.7$ it is $8$ (right).

@@

The \emph{inverse} problem is, given $\tiles \in \mathbb N$, to determine the infimum length of all segments that touch at least $\tiles$ tiles. For real-valued lengths this infimum is not a minimum, because given a segment that touches $\tiles$ tiles, its length can always be reduced by some non-zero amount without changing the number of touched tiles. This is a consequence of the interior of the each tile being an open set.

@@

These problems are related to \ldots ***maybe a short paragraph here mentioning related problems: Buffon's needle, Buffon-Laplace problem. Some references: ``Buffon's Noodle Problem'', J.F.~Ramaley, 1969. ``On Laplace's Extension of the Buffon Needle Problem'', B.J.~Arnow, 1994. \\
\url{https://www.jstor.org/stable/2317945} \\
\url{https://www.jstor.org/stable/2687085}

This paper analyzes both problems for real-valued lengths, and for integer-valued lengths on a unit square grid. Section~\ref{part: fund results} presents fundamental results, from which the characterization is obtained in Section~\ref{part: arbitrary grid, real lengths} for an arbitrary grid with real-valued lengths; in Section~\ref{part: square grid, real lengths} for a square grid with real-valued lengths; and in Section~\ref{part: unit square grid, integer lengths} for a unit square grid with integer-valued lengths.

@@

Knowing a minimal sufficient set of $(i,j)$ pairs facilitates the solution of both the direct an inverse problems, because it reduces the number of pairs that need to be tried. In order to derive a general method to build a minimal sufficient set, it is insightful to consider two specific cases first, as will be presented in the following. It is also helpful to use the equivalent formulation of the direct problem referred to in Section~\ref{part: intro}, i.e.\ finding the maximum number of tiles touched by segments of length $\len$ \emph{or less}.

@@

\section{Arbitrary grid with real-valued lengths}
\label{part: arbitrary grid, real lengths}

For a grid with parameters $a, b \in \mathbb R^+$, the maximum number $\tiles$ of touched tiles for a given real-valued length $\len$ can be represented by a function $\funt: \mathbb R^+ \to \mathbb N$ such that $\tiles = \funt(\len)$. Similarly, for the inverse problem a function $\funl: \mathbb N \to \mathbb R^+$ can be defined such that $\funl(\tiles)$ gives the infimum length of all segments that touch at least $\tiles$ tiles.

@@

Theorems~\ref{theo: funt, form} and \ref{theo: funl, form} not only give the solutions $\funt(\len)$ and $\funl(\tiles)$ to the two problems stated in Section~\ref{part: intro}; they also provide a way to actually position a segment of length slightly greater than $\len$ or $\funl(\tiles)$ so that it touches $\tiles$ or $\funt(\len)$ tiles. Namely, the segment should have its endpoints respectively in the interior of two tiles separated $\isoli-1$ steps horizontally and $\jsoli-1$ steps vertically, with its exact position and orientation  adjusted to avoid any grid points.

@@

For $a=b$, either from symmetry considerations or particularizing the formulas in this section it stems that the optimal orientation of the segment is close to $45^\circ$. This case will be dealt with in Section~\ref{part: square grid, real lengths}, as it lends itself to somewhat simplified formulas.

@@

\section{Square grid with real-valued lengths}
\label{part: square grid, real lengths}

Particularizing the results in Section~\ref{part: arbitrary grid, real lengths} to a square grid, $a=b$, obviously yields simpler formulas.

@@

\section{Unit square grid with integer lengths}
\label{part: unit square grid, integer lengths}

A natural variation of the two problems introduced in Section~\ref{part: intro} is to consider $a=b=1$, with the additional restriction that the segment length can only be a positive integer (equivalently, a square grid is considered with step $a$ and the segment length can only be an integer multiple of $a$).

The \emph{direct problem} in this setting corresponds to the restriction of the function $\funt$ to $\mathbb N$. This will be denoted as a function $\funti: \mathbb N \to \mathbb N$ for greater clarity, although obviously $\funti(\leni) = \funt(\leni)$ for all $\leni \in \mathbb N$. The sequence $\funti(\leni)$, $\leni \in \mathbb N$ takes values $3, 5, 7, 8, 9, \ldots$, and is depicted in Figure~\ref{fig: funti}. This is A346232 in the On-Line Encyclopedia of Integer Sequences \cite{OEIS_unitsq_int_dir}. For this sequence, a simpler expression can be obtained than those resulting from Section~\ref{part: square grid, real lengths} with $a=1$, and the following properties hold.

@@

>>>>

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gices infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\mss}{M}

@@

\title{
On the number of tiles touched by a\\
line segment on a rectangular grid
}

% Authors with different affiliations: from https://tex.stackexchange.com/a/370813/90222
\author[1]{Luis Mendo}
\author[2]{Alex Arkhipov}

@@

Given $a, b \in \mathbb R^+$, consider a grid on $\mathbb R^2$ formed by rectangular \emph{tiles} of width $a$ and height $b$. A line segment of length $\len \in \mathbb R^+$ is located on the plane with arbitrary position and orientation. The segment is said to \emph{touch} a tile if it intersects its interior.\footnote{
The definition uses the interior of the tile, excluding the border, to avoid uninteresting results such as a ``zero-length'' segment touching (a vertex of) $4$ tiles.} This paper deals with the following problems:
\begin{itemize}
\item What is the maximum number of tiles that the segment can touch?
\item What length should a segment have to touch a given number of tiles?
\item What is the average number of tiles touched by a uniformly random segment of a given length?
\end{itemize}

As an example of the first problem, consider $a=1.35$, $b=1$. A segment of unit length can be placed as shown in Figure~\ref{fig: examples} (left) to touch $3$ tiles. In fact, the solution for $\len=1$ is $3$ tiles. The figure also illustrates that the solution for length $2.4$ is $5$ (center), and for $4.7$ it is $8$ (right).

@@

The \emph{inverse} problem is, given $\tiles \in \mathbb N$, to determine the infimum length of all segments that touch at least $\tiles$ tiles. For real-valued lengths this infimum is not a minimum, because given a segment that touches $\tiles$ tiles, its length can always be reduced by some non-zero amount without changing the number of touched tiles. This is a consequence of the interior of each tile being an open set.

The direct and inverse problems are closely related. Namely, if $\len$ is the infimum of all lengths that allow touching at least $\tiles$ tiles (inverse problem), $\tiles$ is the maximum number of tiles that can be touched with lengths slightly greater than $\len$ (direct problem).

The question of determining how many tiles a segment of a given length touches on average will be referred to as the \emph{average-number} problem. As will be seen, solving this also answers the inverse question (segment length to touch a given number of tiles on average), which thus need not be considered separately. By symmetry, one endpoint of the segment can be assumed to lie in a fixed, reference tile. The position of this endpoint will be considered to be \emph{uniformly} distributed on the tile. The segment orientation will be have a \emph{uniform} distribution on the set of all possible directions, and is \emph{independent} of the endpoint position.

@@

The problems addressed in this paper are related to \ldots ***maybe a short paragraph here mentioning related problems: Buffon's needle, Buffon-Laplace problem. Some references: ``Buffon's Noodle Problem'', J.F.~Ramaley, 1969. ``On Laplace's Extension of the Buffon Needle Problem'', B.J.~Arnow, 1994. \\
\url{https://www.jstor.org/stable/2317945} \\
\url{https://www.jstor.org/stable/2687085}

The rest of the paper is organized as follows. \S\ref{part: fund results} presents fundamental results, which form the basis of the subsequent analysis. The direct and inverse problems are considered in \S\ref{part: max}, first for real-valued lengths, and then for integer-valued lengths on a unit square grid. The average-number problem is analyzed in \S\ref{part: ave}.

@@

Knowing a minimal sufficient set of $(i,j)$ pairs facilitates the solution of both the direct an inverse problems, because it reduces the number of pairs that need to be tried. In order to derive a general method to build a minimal sufficient set, it is insightful to consider two specific cases first, as will be presented in the following. It is also helpful to use the equivalent formulation of the direct problem referred to in \S\ref{part: intro}, i.e.\ finding the maximum number of tiles touched by segments of length $\len$ \emph{or less}.

@@

\section{Maximum number of tiles: direct and inverse problems}
\label{part: max}
% ***Is it clear enough that "maximum number of tiles" can refer to the inverse problem too? (because the inverse problem is closely related to the direct one, as discussed in the introduction)
% If not, an alernative would be a more explicit title like: "Maximum number of tiles fo a given length. Conditions on length for a given number of tiles"

The direct and inverse problems for the maximum number of tiles are considered in this section. The general case for rectangular grids with real-valued segment lengths is analyzed first, in \S\ref{part: max: arbitrary grid, real lengths}. The particular case of square grids is addressed in \S\ref{part: max: square grid, real lengths}, as it allows a specialized formula for the direct problem. Lastly, the analysis of a unit square grid with integer-valued lengths is given in \S\ref{part: max: unit square grid, integer lengths}.

\subsection{Arbitrary grid with real-valued lengths}
\label{part: max: arbitrary grid, real lengths}

For a grid with parameters $a, b \in \mathbb R^+$, the maximum number $\tiles$ of touched tiles with a given real-valued length $\len$ can be represented by a function $\funt: \mathbb R^+ \to \mathbb N$ such that $\tiles = \funt(\len)$. Similarly, for the inverse problem a function $\funl: \mathbb N \to \mathbb R^+$ can be defined such that $\funl(\tiles)$ gives the infimum length of all segments that touch at least $\tiles$ tiles.

@@

Theorems~\ref{theo: funt, form} and \ref{theo: funl, form} not only give the solutions $\funt(\len)$ and $\funl(\tiles)$ to the two problems stated in \S\ref{part: intro}; they also provide a way to actually position a segment of length slightly greater than $\len$ or $\funl(\tiles)$ so that it touches $\tiles$ or $\funt(\len)$ tiles. Namely, the segment should have its endpoints respectively in the interior of two tiles separated $\isoli-1$ steps horizontally and $\jsoli-1$ steps vertically, with its exact position and orientation  adjusted to avoid any grid points.

@@

For $a=b$, either from symmetry considerations or particularizing the formulas in the previous theorems it stems that the optimal orientation of the segment is close to $45^\circ$. This case will be dealt with in \S\ref{part: max: square grid, real lengths}, as it lends itself to somewhat simplified formulas.

@@

\subsection{Square grid with real-valued lengths}
\label{part: max: square grid, real lengths}

Particularizing the results in \S\ref{part: max: arbitrary grid, real lengths} to a square grid, $a=b$, obviously yields simpler formulas.

@@

\subsection{Unit square grid with integer lengths}
\label{part: max: unit square grid, integer lengths}

A natural variation of the two problems introduced in \S\ref{part: intro} is to consider $a=b=1$, with the additional restriction that the segment length can only be a positive integer (equivalently, a square grid is considered with step $a$ and the segment length can only be an integer multiple of $a$).

The \emph{direct problem} in this setting corresponds to the restriction of the function $\funt$ to $\mathbb N$. This will be denoted as a function $\funti: \mathbb N \to \mathbb N$ for greater clarity, although obviously $\funti(\leni) = \funt(\leni)$ for all $\leni \in \mathbb N$. The sequence $\funti(\leni)$, $\leni \in \mathbb N$ takes values $3, 5, 7, 8, 9, \ldots$, and is depicted in Figure~\ref{fig: funti}. This is A346232 in the On-Line Encyclopedia of Integer Sequences \cite{OEIS_unitsq_int_dir}. For this sequence, a simpler expression can be obtained than those resulting from \S\ref{part: max: square grid, real lengths} with $a=1$, and the following properties hold.

@@

\section{Average number of tiles}
\label{part: ave}

***A theorem, and maybe one or two propositions. It is interesting to state $\Pr[i \geq i_0]$ (averaged over $x$) in a proposition, as it can be compared with the Buffon experiment.

***Define the average number of tiles touched as a function $\funta: \mathbb R^+ \to \mathbb R^+$.

***The formula for the average number of files is:
\begin{equation}
\funta(\len) = \frac{2\len}{\pi}\left(\frac 1 a + \frac 1 b\right) + 1.
\end{equation}

***Explain that the inverse question is straightforward, because $\funta$ is monotone increasing, thus it has an inverse function (defined on the range of $\funta$). In fact, as will be seen, $\funta$ is an affine function, and therefore it is immediate to obtain its inverse.

***Some mention that there is no need to study square grids or the case of integer-valued lengths separately.

***Relation the the problem of the Buffon needle. On one hand, the formula for b=infinity gives the average number of lines crossed in the Buffon experiment. On the other hand, the intermediate values $\Pr[i \geq i_0]$ (proposition) are a generalization of the probability of crossing at least a line in the Buffon experiment, when the needle endpoint is only allowed to move in a smaller vertical strip.

***Mention particular values: a=b=1 gives...; a=b=4/pi gives len+1. Can we find any ``meaning'' for these particular cases?

***The function is affine, not linear. This is related to the fact that the parts of the segment near the endpoints can touch several tiles

***Maybe a graph comparing maximum and average. Or compute/plot their ratio. The asymptotic slopes are $\sqrt(1/a^2+1/b^2)$ for $\funt$ and $2/\pi(1/a+1/b)$. It may be interesting to consider the ratio of asymptotic slopes. Some quick plot of that ratio suggests it is maximum for $a=b$, and equals $sqrt(8)/pi = 0.9003$.

***The distribution would be interesting, not just the average. But it is more difficult to obtain, because of the statistical relationship between the random variables i and j. This relationship does not affect the \emph{average} of the random variable $i+j-1$ (for a given length), but it does affect the its \emph{distribution}. [I think it is interesting to mention this].

\begin{theorem}
\label{theo: funta, form}
\end{theorem}
